---
title: "Week 5 In-Class Lab: Where the Model Is Wrong"
subtitle: "MUSA 5080 · Linear Regression for Prediction"
format:
  html:
    toc: true
    code-tools: true
execute:
  warning: false
  message: false
freeze: false
---

The first part of the lab (Parts 0–7) is guided and builds your toolkit. **Type the code yourself**; don't copy and paste. Fill in every `___`. The last part is the **challenge**: your team builds the best rent prediction model it can and defends it. At the end, the presenter from each team is chosen at random, so everyone needs to understand every step.

> **The question:** A housing agency wants an *expected* rent for every Philadelphia tract, so it can flag tracts where rents run well above what we'd expect given who lives there.
>
> Is that a β̂ problem or a ŷ problem? Write your answer here before you start: **A ŷ problem.** The agency needs a good *predicted* rent for every tract to compare against the actual rent; the individual coefficients are not the deliverable.

---

## Part 0: Setup

```{r}
#| label: setup
library(tidyverse)
library(tidycensus)
library(sf)
library(caret)     # cross-validation
library(lmtest)    # Breusch-Pagan test

# Your key lives in .Renviron. Never paste it here.
Sys.getenv("CENSUS_API_KEY") != ""   # should print TRUE
```

---

## Part 1: Rebuild the table you already know

This is last week's pull with a few more variables. Every line should look familiar.

```{r}
#| label: pull-tracts
philly_tracts <- get_acs(
  geography = "tract",
  state = "PA",
  county = "Philadelphia",
  variables = c(
    med_rent   = "B25064_001",   # median gross rent (our Y)
    med_inc    = "B19013_001",   # median household income
    total_pop  = "B01003_001",
    occ_units  = "B25003_001",   # occupied housing units
    renter_occ = "B25003_003",   # renter-occupied units
    edu_total  = "B15003_001",   # population 25+
    ba         = "B15003_022",
    ma         = "B15003_023",
    prof       = "B15003_024",
    phd        = "B15003_025"
  ),
  year = 2023,
  output = "wide",
  geometry = TRUE
) %>%
  st_transform(2272)          # Pennsylvania South State Plane, feet
```

Now build the variables. **A data frame goes in, a data frame comes out.**

```{r}
#| label: build-vars
philly_tracts <- philly_tracts %>%
  mutate(
    pct_renter = renter_occE / occ_unitsE * 100,
    pct_ba     = (baE + maE + profE + phdE) / edu_totalE * 100,
    rent_se    = med_rentM / 1.645,
    rent_cv    = rent_se / med_rentE * 100
  )
```

Bring back the stations and count how many sit within half a mile of each tract. This is Week 4 with one new function.

```{r}
#| label: stations
stations <- st_read("https://www.rideindego.com/stations/json/", quiet = TRUE) %>%
  st_transform(st_crs(philly_tracts))

philly_tracts <- philly_tracts %>%
  mutate(n_stations = lengths(st_is_within_distance(., stations, dist = 2640)))   # half a mile = 2,640 ft
```

**Stop and look:** `st_is_within_distance()` returns a list, one element per tract, holding the stations within range. What does `lengths()` do to that list?

### Which tracts can we model?

A regression can't use a row with a missing Y or X. R will quietly drop those rows, and later your residuals won't line up with your tracts. So drop them **yourself**, on purpose, and count.

```{r}
#| label: filter
nrow(philly_tracts)

tracts <- philly_tracts %>%
  filter(total_popE > 0,
         !is.na(med_rentE),
         !is.na(med_incE),
         !is.na(pct_ba),
         !is.na(pct_renter))

nrow(tracts)
```

**Rule from Week 4, again:** how many tracts did you lose? Which kind of tract has no median rent? (Hint: think about who lives there.)

---

## Part 2: Look before you model

The Week 3 habit comes first: plot the relationship before you fit anything.

```{r}
#| label: scatter
ggplot(tracts, aes(x = med_incE, y = med_rentE)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  scale_x_continuous(labels = scales::dollar) +
  scale_y_continuous(labels = scales::dollar) +
  labs(x = "Median household income", y = "Median gross rent",
       title = "Rent and income, Philadelphia tracts") +
  theme_minimal()
```

**Stop and look:** Does a straight line describe this cloud well everywhere, or does it bend? Where are the points that sit far from the line?

---

## Part 3: Fit the line

```{r}
#| label: m1
m1 <- lm(med_rentE ~ med_incE, data = tracts)
summary(m1)
```

**Units first.** Translate the income coefficient into something a person would notice:

```{r}
#| label: translate
coef(m1)["med_incE"] * 10000    # predicted rent change for +$10,000 of income
```

Write one sentence: "A tract with $10,000 more in median income has a predicted rent about $`r round(coef(m1)["med_incE"] * 10000)` higher per month."

**Stop and look:** The p-value is tiny. Does that tell you whether $`r round(coef(m1)["med_incE"] * 10000)`/month is a big effect? What would you need to know to decide?

---

## Part 4: Read the same output twice

Add the other predictors.

```{r}
#| label: m2
m2 <- lm(med_rentE ~ med_incE + pct_renter + pct_ba + n_stations, data = tracts)
summary(m2)
```

Look at the `n_stations` coefficient. Answer both questions in one or two sentences each:

1. **β̂ reading:** Can you tell the housing agency that bike share stations raise rents? Why or why not?

   *Your answer:* No. This is an observational cross-section. Indego sites stations in dense, central, high-demand neighborhoods, so `n_stations` stands in for centrality and amenities the model doesn't measure (omitted variables), and demand may drive station placement rather than the reverse. The coefficient is an association, not a causal effect.

2. **ŷ reading:** Should `n_stations` stay in a model whose only job is predicting rent? Why or why not?

   *Your answer:* Yes, if it lowers out-of-sample error (check the CV RMSE in Part 6). The agency has station counts for every tract, the variable is not built from rent, and being a proxy for location desirability is exactly what makes it useful for prediction.

::: {.callout-note}
## Same number, two questions
If your answers to 1 and 2 differ, that's the point. For inference, a variable that stands in for something unmeasured is a problem. For prediction, it can be exactly why the model works.
:::

One more check. Compare in-sample fit:

```{r}
#| label: r2
summary(m1)$r.squared
summary(m2)$r.squared
```

Hold on to these numbers. Part 6 asks whether they mean what they seem to.

---

## Part 5: Uncertainty in input data

Our Y is an ACS estimate with its own margin of error. Add the residuals to the table and check whether the biggest misses happen where the estimate is least reliable.

```{r}
#| label: fog
tracts <- tracts %>%
  mutate(fitted_m2 = fitted(m2),
         resid_m2  = residuals(m2))

ggplot(tracts, aes(x = rent_cv, y = abs(resid_m2))) +
  geom_point(alpha = 0.5) +
  geom_smooth(se = FALSE) +
  labs(x = "CV of the median rent estimate (%)",
       y = "Size of the model's miss ($)",
       title = "Are the biggest misses in the noisiest tracts?") +
  theme_minimal()
```

**Stop and look:** If a tract's rent estimate has a CV of 35%, how much of the model's miss there could be survey noise rather than a model failure? What does that mean for flagging "tracts with rents above expectation"?

---

## Part 6: Fit vs. prediction

::: {.callout-warning}
## Two different CVs
**Week 3 CV** = coefficient of variation: how noisy is *one estimate*?
**Today's CV** = cross-validation: how well does a *model* predict data it hasn't seen?
Same letters, unrelated ideas.
:::

### 6A: One train/test split

```{r}
#| label: split
set.seed(5080)
n <- nrow(tracts)
train_idx <- sample(1:n, size = 0.8 * n)

train_data <- tracts[train_idx, ]
test_data  <- tracts[-train_idx, ]

m2_train <- lm(med_rentE ~ med_incE + pct_renter + pct_ba + n_stations,
               data = train_data)

test_pred <- predict(m2_train, newdata = test_data)

rmse_train <- sqrt(mean(residuals(m2_train)^2))
rmse_test  <- sqrt(mean((test_data$med_rentE - test_pred)^2))

c(train = rmse_train, test = rmse_test)
```

**Stop and look:** Which RMSE is larger? Why would you expect that? In words: "The model's typical miss on tracts it hasn't seen is about $`r round(rmse_test)` per month."

Change the seed to any other number and rerun the chunk. Did the test RMSE change? That's why one split isn't enough.

### 6B: 10-fold cross-validation

`caret` wants a plain data frame, so drop the geometry first. (The geometry isn't a predictor, and `train()` will choke on it.)

```{r}
#| label: cv
model_df <- tracts %>% st_drop_geometry()

ctrl <- trainControl(method = "cv", number = 10)

set.seed(5080)
cv_m1 <- train(med_rentE ~ med_incE,
               data = model_df, method = "lm", trControl = ctrl)

set.seed(5080)
cv_m2 <- train(med_rentE ~ med_incE + pct_renter + pct_ba + n_stations,
               data = model_df, method = "lm", trControl = ctrl)

cv_m1$results
cv_m2$results
```

Setting the same seed before each `train()` gives both models the same folds, so the comparison is fair.

**Stop and look:** Fill in the table.

| | In-sample R² | CV RMSE |
|---|---|---|
| m1 | `r round(summary(m1)$r.squared, 3)` | $`r round(cv_m1$results$RMSE, 1)` |
| m2 | `r round(summary(m2)$r.squared, 3)` | $`r round(cv_m2$results$RMSE, 1)` |

Did adding variables improve R²? Did it improve CV RMSE by as much? For a prediction task, which column is the grade?

---

## Part 7: Where is the model wrong?

### 7A: Residuals vs. fitted

```{r}
#| label: resid-fitted
ggplot(tracts, aes(x = fitted_m2, y = resid_m2)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, color = "red") +
  labs(x = "Predicted rent", y = "Residual (actual − predicted)") +
  theme_minimal()

bptest(m2)
```

**Stop and look:** Random cloud, funnel, or curve? If it's a funnel, the predictions aren't biased, but one RMSE hides the fact that the model misses by much more in some tracts. Which tracts?

### 7B: The map

The residuals are already a column in an sf data frame. Mapping them is one `geom_sf()` away.

```{r}
#| label: resid-map
ggplot(tracts) +
  geom_sf(aes(fill = resid_m2), color = NA) +
  scale_fill_gradient2(low = "#2166ac", mid = "grey95", high = "#b2182b",
                       midpoint = 0, labels = scales::dollar) +
  theme_void() +
  labs(title = "Where the rent model is wrong",
       subtitle = "Red = rent higher than predicted · Blue = lower than predicted",
       fill = "Residual")
```

**Stop and look:** If the errors were random, the map would look speckled. Does it? Where do the reds sit together? What might those places have in common that the model doesn't know about?

Write down your answer. We'll come back to it in Weeks 6 and 7.

---

## Part 8: The Challenge

Your team's job: build the **best model for predicting tract median rent**, then use it to answer the agency's question.

### Rules

1. **The grade is 10-fold CV RMSE.** Use `model_df` (or `tracts`) and run `set.seed(5080)` directly before `train()`, exactly as in Part 6B, so every team is tested on the same folds.
2. **Don't drop rows.** If you add a variable with missing values, you change the folds and the comparison stops being fair. Fill the gaps or choose a different variable.
3. **You may add** any ACS variables (use `load_variables(2023, "acs5")` to browse) and any spatial feature you can build with Week 4 tools: distances, buffers, counts.
4. **You may not use anything built from rent itself.** That rules out `med_rentM`, `rent_cv`, rent burden (`B25071`), and the rent distribution tables (`B25063`). Ask yourself: would the agency *have* this variable for a tract whose rent it's trying to predict? If not, using it is cheating the test. That's called **leakage**.

### Deliver

1. **Your CV RMSE**, in dollars, written as a sentence.
2. **One feature you added, and why.** What did you think it would capture that the census variables miss? Did it help?
3. **One coefficient, read twice.** Pick a variable in your final model and give its β̂ reading and its ŷ reading in one sentence each.
4. **The agency's list.** Using your final model fit to all tracts, find the five tracts with the largest *positive* residuals (rent well above expectation). For each, report the residual and the rent CV. Which of the five would you tell the agency to trust, and which might be fog?
5. **Your residual map.** Did your added features make the clusters from Part 7B shrink?

#### 8.1 Extra ACS variables (none built from rent)

```{r}
#| label: challenge-acs
extra_acs <- get_acs(
  geography = "tract",
  state = "PA",
  county = "Philadelphia",
  variables = c(
    med_value   = "B25077_001",   # median value, owner-occupied units
    yr_built    = "B25035_001",   # median year structure built
    med_rooms   = "B25018_001",   # median number of rooms
    hu_total    = "B25002_001",   # housing units
    hu_vacant   = "B25002_003",   # vacant units
    pov_univ    = "B17001_001",   # poverty universe
    pov_below   = "B17001_002",   # below poverty
    workers     = "B08301_001",   # workers 16+
    transit     = "B08301_010",   # commute by public transit
    units_total = "B25024_001",   # units in structure: total
    u_1det      = "B25024_002",   # 1-unit detached
    u_1att      = "B25024_003",   # 1-unit attached (rowhomes)
    u_20_49     = "B25024_008",   # 20-49 units
    u_50plus    = "B25024_009",   # 50+ units
    med_age     = "B01002_001"    # median age
  ),
  year = 2023,
  output = "wide",
  geometry = FALSE
) %>%
  transmute(
    GEOID,
    med_value   = med_valueE,
    yr_built    = yr_builtE,
    med_rooms   = med_roomsE,
    pct_vacant  = hu_vacantE / hu_totalE * 100,
    pct_poverty = pov_belowE / pov_univE * 100,
    pct_transit = transitE / workersE * 100,
    pct_single  = (u_1detE + u_1attE) / units_totalE * 100,
    pct_bigbldg = (u_20_49E + u_50plusE) / units_totalE * 100,
    med_age     = med_ageE
  )
```

#### 8.2 Spatial features (Week 4 tools)

```{r}
#| label: challenge-spatial
city_hall <- st_sfc(st_point(c(-75.1636, 39.9526)), crs = 4326) %>%
  st_transform(st_crs(tracts))

tract_pts <- st_centroid(st_geometry(tracts))
nearest   <- st_nearest_feature(tract_pts, stations)

tracts_ch <- tracts %>%
  mutate(
    dist_ch_mi      = as.numeric(st_distance(tract_pts, city_hall)[, 1]) / 5280,
    dist_station_mi = as.numeric(st_distance(tract_pts, stations[nearest, ],
                                             by_element = TRUE)) / 5280
  ) %>%
  left_join(extra_acs, by = "GEOID")   # left_join keeps every row, in the same order
```

#### 8.3 Fill the gaps; don't drop rows (Rule 2)

```{r}
#| label: challenge-impute
impute_vars <- c("med_value", "yr_built", "med_rooms", "pct_vacant", "pct_poverty",
                 "pct_transit", "pct_single", "pct_bigbldg", "med_age")

# How many gaps are there before filling?
colSums(is.na(st_drop_geometry(tracts_ch)[impute_vars]))

tracts_ch <- tracts_ch %>%
  mutate(
    value_missing = as.integer(is.na(med_value)),   # flag tracts with no home-value estimate
    across(all_of(impute_vars), ~ if_else(is.na(.x), median(.x, na.rm = TRUE), .x))
  )

# Same rows, same order as Part 6B, so the folds are identical
stopifnot(nrow(tracts_ch) == nrow(tracts),
          identical(tracts_ch$GEOID, tracts$GEOID))

model_df_ch <- tracts_ch %>% st_drop_geometry()
```

#### 8.4 Compare candidate models with 10-fold CV

```{r}
#| label: challenge-cv
f_base  <- med_rentE ~ med_incE + pct_renter + pct_ba + n_stations
f_value <- update(f_base, . ~ . + log(med_value) + value_missing)
f_a     <- med_rentE ~ log(med_incE) + pct_renter + pct_ba + n_stations +
             log(med_value) + value_missing + med_rooms + yr_built
f_b     <- update(f_a, . ~ . + pct_vacant + pct_poverty + pct_transit +
                    pct_single + pct_bigbldg + med_age +
                    log1p(dist_ch_mi) + log1p(dist_station_mi))

set.seed(5080)
cv_base  <- train(f_base,  data = model_df_ch, method = "lm", trControl = ctrl)
set.seed(5080)
cv_value <- train(f_value, data = model_df_ch, method = "lm", trControl = ctrl)
set.seed(5080)
cv_a     <- train(f_a,     data = model_df_ch, method = "lm", trControl = ctrl)
set.seed(5080)
cv_b     <- train(f_b,     data = model_df_ch, method = "lm", trControl = ctrl)

cv_results <- tibble(
  model = c("Base (m2)", "Base + home value", "A: + log income, housing stock",
            "B: A + neighborhood & location"),
  RMSE     = c(cv_base$results$RMSE, cv_value$results$RMSE,
               cv_a$results$RMSE,    cv_b$results$RMSE),
  Rsquared = c(cv_base$results$Rsquared, cv_value$results$Rsquared,
               cv_a$results$Rsquared,    cv_b$results$Rsquared),
  MAE      = c(cv_base$results$MAE, cv_value$results$MAE,
               cv_a$results$MAE,    cv_b$results$MAE)
)
knitr::kable(cv_results, digits = c(0, 1, 3, 1))

best_i        <- which.min(cv_results$RMSE)
best_name     <- cv_results$model[best_i]
final_formula <- list(f_base, f_value, f_a, f_b)[[best_i]]
```

#### 8.5 Final model on all tracts and the agency's list

```{r}
#| label: challenge-final
m_final <- lm(final_formula, data = tracts_ch)
summary(m_final)

tracts_ch <- tracts_ch %>%
  mutate(fitted_final = fitted(m_final),
         resid_final  = residuals(m_final))

top5 <- tracts_ch %>%
  st_drop_geometry() %>%
  arrange(desc(resid_final)) %>%
  slice_head(n = 5) %>%
  transmute(
    GEOID, NAME,
    actual    = med_rentE,
    predicted = round(fitted_final),
    residual  = round(resid_final),
    rent_moe  = med_rentM,
    rent_cv   = round(rent_cv, 1),
    reliability = case_when(rent_cv < 12  ~ "High (CV < 12%)",
                            rent_cv <= 40 ~ "Moderate (12-40%)",
                            TRUE          ~ "Low (CV > 40%)"),
    verdict = case_when(
      resid_final > med_rentM & rent_cv < 12 ~ "Trust",
      resid_final > med_rentM                ~ "Probably real; verify",
      TRUE                                   ~ "Could be fog"
    )
  )
knitr::kable(top5)
```

A residual larger than the rent's 90% margin of error (`rent_moe`) can't be explained by sampling noise alone. A residual smaller than the MOE could be fog.

#### 8.6 Residual map: Part 7B vs. final model

```{r}
#| label: challenge-map
#| fig-width: 10
#| fig-height: 6
resid_long <- tracts_ch %>%
  select(resid_m2, resid_final) %>%
  pivot_longer(c(resid_m2, resid_final), names_to = "model", values_to = "residual") %>%
  mutate(model = recode(model, resid_m2 = "Part 7B model (m2)",
                               resid_final = "Our final model"),
         model = factor(model, levels = c("Part 7B model (m2)", "Our final model")))

ggplot(resid_long) +
  geom_sf(aes(fill = residual), color = NA) +
  scale_fill_gradient2(low = "#2166ac", mid = "grey95", high = "#b2182b",
                       midpoint = 0, labels = scales::dollar) +
  facet_wrap(~ model) +
  theme_void() +
  labs(title = "Where the rent model is wrong, before and after",
       subtitle = "Red = rent higher than predicted · Blue = lower than predicted",
       fill = "Residual")

c(sd_m2 = sd(tracts_ch$resid_m2), sd_final = sd(tracts_ch$resid_final))
```

### Our answers

1. **CV RMSE.** Our best model (`r best_name`) has a 10-fold CV RMSE of about $`r round(min(cv_results$RMSE))`. On tracts it hasn't seen, its typical miss is about $`r round(min(cv_results$RMSE))` per month, down from $`r round(cv_base$results$RMSE)` for the Part 6 model m2.

2. **One feature we added: median home value (`B25077`).** Income describes who lives in a tract *now*, but home values capture what the location and housing stock are worth: amenities, housing quality, and gentrification pressure that current residents' incomes may not show yet. It is not leakage: it comes from owners, not rents, and the agency would have it for any tract. Adding only home value (plus a missing flag) to m2 moves CV RMSE from $`r round(cv_base$results$RMSE)` to $`r round(cv_value$results$RMSE)`.

3. **One coefficient, read twice: `pct_ba`.** In the final model, `pct_ba` has a coefficient of `r round(coef(m_final)["pct_ba"], 2)`, so a tract with 10 more percentage points of adults holding a BA or higher has a predicted rent about $`r round(coef(m_final)["pct_ba"] * 10)` different.
   - *β̂ reading:* Holding the other variables fixed, more educated tracts are *associated* with that difference in rent. This does not mean degrees raise rents; education stands in for neighborhood demand and amenities we don't measure.
   - *ŷ reading:* For prediction, `pct_ba` stays if it helps CV RMSE. Whether it is causal doesn't matter, only whether it carries information about rent the other variables don't.

4. **The agency's list.** The table in 8.5 gives the five tracts with the largest positive residuals, with each residual, MOE, and rent CV. Tracts to trust: `r paste(top5$GEOID[top5$verdict == "Trust"], collapse = ", ")`. Tracts that might be fog: `r paste(top5$GEOID[top5$verdict == "Could be fog"], collapse = ", ")`. The rule: we trust a flag only when the residual is bigger than the rent estimate's 90% MOE *and* the rent CV is under 12% (Week 3). If the residual is inside the MOE, the "above expectation" rent could be survey noise, so the agency should verify before acting.

5. **Residual map.** The residual SD falls from $`r round(sd(tracts_ch$resid_m2))` (m2) to $`r round(sd(tracts_ch$resid_final))` (final model). [After viewing the map in 8.6, confirm: did the red clusters from Part 7B (e.g., Center City and the surrounding high-rent areas) shrink or fade? Which clusters remain?]

### Scoring (for bragging rights, not grades)

| | Points |
|---|---|
| Lowest CV RMSE in the room | 3 |
| Second and third lowest | 2 / 1 |
| Feature explanation is clear and specific | 2 |
| Coefficient read correctly both ways | 2 |
| Agency list accounts for reliability (Week 3) | 2 |
| Randomly chosen presenter explains the model clearly | 3 |
| Breaks a rule (leakage or dropped rows) | −5 |

### Before you present, discuss

- One of your variables is not statistically significant, but removing it makes CV RMSE worse. Do you keep it? What would your Spatial Statistics instructor say, and why is neither of you wrong?
- A council member sees your model and says, "So your model proves that renters drive up rents." What do you say?
- If the agency uses your top-five list to target inspections, whose rents does a model error hurt most?

---

## If you're curious (ungraded, no extra points)

- Replace `med_incE` with `log(med_incE)`. Does the curve in Part 2 straighten? Does CV RMSE improve? How do you read the new coefficient?
- Try a random forest: change `method = "lm"` to `method = "ranger"` (run `install.packages("ranger")` first). Does it beat your linear model? By how much? Is the gain worth losing the ability to explain the model to the agency?
- Your residual map probably still has clusters. Build one spatial feature that you think would absorb them, add it, and remap. That's next week in miniature.
