---
title: "Assignment 2: Who Do We Know the Least About?"
subtitle: "Measuring uncertainty in American Community Survey data"
author: "Rutong Chen"
date: today
format:
  html:
    toc: true
    code-fold: false
execute:
  warning: false
  message: false
---

## Setup

```{r}
library(tidyverse)
library(tidycensus)
```

---

## The question

Every number published by the American Community Survey is an estimate based on
a sample rather than a complete count. Each estimate therefore comes with a
margin of error that describes its statistical uncertainty.

In this analysis, I examine median gross rent across counties in New York State
and ask which communities are measured most and least reliably. Understanding
the uncertainty in rent estimates is important because these data may influence
housing policy and the allocation of affordable housing resources.

---

## Choosing what to measure

```{r}
my_state <- "NY"
my_variable <- "B25064_001"
```

## Pulling the data

```{r}
county_data <- get_acs(
  geography = "county",
  variables = my_variable,
  state = my_state,
  year = 2023,
  survey = "acs5"
)
```

Let's look at what came back.

```{r}
dim(county_data)
head(county_data)
```

The dataset contains 62 rows, with one observation for each county-level unit in
New York State. This matches the 62 counties and county equivalents in the
state.

---

## How big is the margin of error?

A margin of error of 100 dollars means something different for an estimate of
800 dollars than for an estimate of 2,000 dollars. To compare uncertainty across
counties, I calculate the margin of error as a percentage of the estimate.

```{r}
county_data <- county_data %>%
  mutate(moe_pct = moe / estimate * 100)
```

Now I sort the counties to identify the least reliable estimates.

```{r}
county_data %>%
  arrange(desc(moe_pct)) %>%
  select(NAME, estimate, moe, moe_pct) %>%
  head(10)
```

Hamilton County has the largest margin of error relative to its estimate. Its
estimated median gross rent is 860 dollars, with a margin of error of 95 dollars.
The margin of error is therefore approximately 11.0 percent of the estimate.

---

## Sorting counties into reliability categories

```{r}
county_data <- county_data %>%
  mutate(reliability = case_when(
    moe_pct < 5  ~ "High confidence",
    moe_pct < 10 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))

count(county_data, reliability)
```

Using these thresholds, 51 counties are classified as high confidence, 10 as
moderate, and one as low confidence. These thresholds are analytical categories
used to compare counties in this report rather than official Census Bureau
standards.

---

## Is the margin of error bigger in small places?

To compare the reliability of the rent estimate with county population, I pull
median gross rent and total population together in wide format.

```{r}
county_wide <- get_acs(
  geography = "county",
  variables = c(myvar = my_variable,
                pop   = "B01003_001"),
  state = my_state,
  year = 2023,
  survey = "acs5",
  output = "wide"
)

head(county_wide)
```

The `E` suffix identifies an estimate, while the `M` suffix identifies its
margin of error. Therefore, `myvarE` and `myvarM` contain the median gross rent
estimate and margin of error, while `popE` and `popM` contain the corresponding
population values.

```{r}
county_wide <- county_wide %>%
  mutate(moe_pct = myvarM / myvarE * 100) %>%
  mutate(reliability = case_when(
    moe_pct < 5  ~ "High confidence",
    moe_pct < 10 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))
```

Now I compare the average population of counties in each reliability category.

```{r}
county_wide %>%
  group_by(reliability) %>%
  summarize(
    n_counties = n(),
    avg_population = mean(popE, na.rm = TRUE)
  )
```

High-confidence counties have an average population of approximately 381,479,
compared with approximately 41,178 in moderate-confidence counties. Hamilton
County, the only low-confidence county, has an estimated population of 5,102.
This pattern suggests that counties with smaller populations tend to have less
precise median gross rent estimates. It is consistent with the smaller ACS
samples generally available in less populated places.

---

## What this means for policy

The results show that Hamilton County is the New York county we know the least
about for median gross rent. Its estimated median gross rent is $860, with a
margin of error of $95, or about 11.0 percent of the estimate. It is the only
county classified as low confidence under the thresholds used here. Schuyler
and Greene Counties also have relatively large margins of error, at
approximately 8.7 percent and 8.0 percent, respectively, and are classified as
moderate confidence.

This pattern is related to the way the American Community Survey works. The ACS
is based on a sample rather than a complete count of every household. Counties
with smaller populations generally provide fewer sampled households, so their
estimates tend to be less precise. In this analysis, high-confidence counties
have an average population of about 381,479, compared with about 41,178 for
moderate-confidence counties and 5,102 for the low-confidence county. These
results are consistent with the expectation that smaller places are measured
with greater uncertainty.

Ignoring this uncertainty could create unfair policy outcomes. If New York
State allocated affordable-housing funds using median rent estimates alone,
Hamilton County could be treated as though its estimate were just as precise as
estimates for much larger counties. A county near a funding threshold might
receive too little assistance, or be excluded entirely, because of sampling
variation rather than a genuine difference in housing needs. Agencies should
therefore consider both the estimate and its margin of error, especially when
making decisions about small rural communities.
